---
title: "ReproducibiliTea"
subtitle: "The Garden of Forking Paths (Gelman and Loken, 2013)"
author: "Clement Lee"
date: "2025-10-16 (Thu)"
output:
  beamer_presentation: default
classoption: aspectratio=169
bibliography: bib/ref-repro.bib
biblio-style: apalike
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
library(dplyr)
library(tibble)
library(ggplot2)
library(igraph)
library(ggraph)
theme_set(theme_bw(12))
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, out.width = "80%")
f <- knitr::include_graphics
```

## Welcome

:::::: {.columns}
::: {.column width="59%" data-latex="{\textwidth}"}

$~$

* A blend of:

  - Open Research
  - https://reproducibilitea.org
  - UK Reproducibility Network (UKRN)

* Monthly informal discussion

* Practices sharing

* Precursor of training

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="39%" data-latex="{\textwidth}"}

```{r logo}
f("images/reproducibiliTea_logo.png")
```

:::
::::::

## Training

:::::: {.columns}
::: {.column width="49%" data-latex="{\textwidth}"}

$$~$$

* Research Software Engineering:

  - https://rse.ncldata.dev/events

* Open Research workshops:

  1. Go to https://elements.ncl.ac.uk
  2. "Find Learning" $\rightarrow$ "Courses"
  3. Search "Open Research"

* Email openresearch@ncl.ac.uk

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="49%" data-latex="{\textwidth}"}

```{r elements}
f("images/elements.png")
```

:::
::::::

## Other events

```{r ora25}
#| out.width: "70%"
f("images/ora25.png")
```

* Open Research Awards (Deadine Nov 7th):

  - https://www.ncl.ac.uk/open-research/awards/

* Statistics & Data Science Clinic:

  - https://www.ncl.ac.uk/maths-physics/engagement/nusolve/

* (External) Replication Games

  - We plan to form a Newcastle team if there's enough interest

## ~~The Garden of Forking Paths~~

:::::: {.columns}
::: {.column width="49%" data-latex="{\textwidth}"}

\begin{align*}
~&~
\end{align*}

* ~~1941 short story by Argentine writer and poet Jorge Luis Borges~~

* ~~On the surface: story of a spy in World War I~~

* ~~Foreshadows the many-worlds interpretation of quantum mechanics~~

* ~~An inspiration of hypertext fiction~~

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="49%" data-latex="{\textwidth}"}

```{r jlb}
f("images/jorge-luis-borges.jpg")
```

:::
::::::

## @gl13 on @bt13

```{r fertility-illustration}
#| fig.asp: 0.5
#| fig.width: 4
df0 <-
  tibble(
    x = seq(-4.0, 4.0, length = 1001),
    y = dnorm(x)
  )
df0 |> ggplot() +
  geom_line(aes(x, y)) +
  scale_x_continuous(breaks = NULL, minor_breaks = NULL) +
  scale_y_continuous(breaks = NULL, minor_breaks = NULL) +
  labs(x = "Day in menstrual cycle", y = "Pr(wear shades\n of red)") +
  theme(
    panel.border = element_blank(),
    axis.title.y = element_text(angle = 0, vjust = 0.5)
  )
```



## Good practice

* Specify the model & criteria before you collect & analyse the data

* Inclusion: Women younger than 40

* Peak: Days 10--17

* Colors: Red $+$ pink

## Illustration

```{r function}
foo <- function(df.edges, dashed = TRUE, black = TRUE) {
  g <- graph_from_data_frame(df.edges, directed = TRUE)
  E(g)$highlight <- df.edges$highlight
  V(g)$label <- V(g)$name
  ggraph(g, layout = "dendrogram") +
    geom_edge_diagonal(
      aes(
        color = highlight,
        linetype = highlight,
        width = highlight
      )
    ) +
    scale_edge_color_manual(values = c(`FALSE` = if (black) "black" else "white", `TRUE` = "red")) +
    scale_edge_linetype_manual(values = c(`FALSE` = if (dashed) "dashed" else "solid", `TRUE` = "solid")) +
    scale_edge_width_manual(values = c(`FALSE` = 0.5, `TRUE` = 2.0)) +
    geom_node_point(size = 4) +
    geom_node_text(aes(label = label), hjust = 0.5, vjust = -1.2, size = 5) +
    coord_flip() +
    scale_y_reverse() +
    theme_void() +
    theme(legend.position = "none")
}
```

```{r no-forking-paths}
#| fig.align: "center"
# Base data frame of edges
edges0 <- tribble(
  ~from, ~to,
  "Start", "Younger than 50",
  "Start", "Younger than 40",
  "Younger than 50", "Days 10-17",
  "Younger than 50", "Days 17-25",
  "Younger than 50", "Days 6-14",
  "Days 10-17", "Red",
  "Days 10-17", "Red + Pink",
  "Days 17-25", "Red ",
  "Days 17-25", "Red + Pink ",
  "Days 6-14", "Red  ",
  "Days 6-14", "Red + Pink  ",
  "Younger than 40", "Days 10-17 ",
  "Younger than 40", "Days 17-25 ",
  "Younger than 40", "Days 6-14 ",
  "Days 10-17 ", "Red   ",
  "Days 10-17 ", "Red + Pink   ",
  "Days 17-25 ", "Red     ",
  "Days 17-25 ", "Red + Pink     ",
  "Days 6-14 ", "Red       ",
  "Days 6-14 ", "Red + Pink       ",
)

# highlight edges
edges1 <- edges0 %>%
  mutate(highlight = case_when(
    (from == "Start" & to == "Younger than 40") |
    (from == "Younger than 40" & to == "Days 10-17 ") |
    (from == "Days 10-17 " & to == "Red + Pink   ") ~ TRUE,
    TRUE ~ FALSE
  ))
foo(edges1, black = FALSE) # dashed doesn't matter
```

## Fishing / $p$-hacking

```{r forking-paths}
#| fig.align: "center"
foo(edges1, dashed = FALSE, black = TRUE)
```

## Rebuttal

Tracy and Beall (2013)

* In a blog post that doesn't exist anymore

* Claimed they were not fishing

* Only one single analysis on one single data set

@gl13

* Even if they were not fishing, they didn't follow the good practice either

* Pointed to **researcher** degrees of freedom

## ~~Researcher~~ degrees of freedom

```{r degrees-of-freedom}
#| fig.align: "center"
#| out.width: "70%"
f("images/degrees-of-freedom.png")
```
Source: https://www.scribbr.co.uk/stats/degrees-of-freedom-df/

## **Researcher** degrees of freedom

```{r forking-paths-highlight}
#| fig.align: "center"
foo(edges1)
```

## Had the data been different

```{r different-data}
#| fig.align: "center"
edges2 <- edges0 %>%
  mutate(highlight = case_when(
    (from == "Start" & to == "Younger than 50") |
    (from == "Younger than 50" & to == "Days 17-25") |
    (from == "Days 17-25" & to == "Red ") ~ TRUE,
    TRUE ~ FALSE
  ))
foo(edges2)
```

## Or this

```{r different-data-again}
#| fig.align: "center"
edges3 <- edges0 %>%
  mutate(highlight = case_when(
    (from == "Start" & to == "Younger than 40") |
    (from == "Younger than 40" & to == "Days 6-14 ") |
    (from == "Days 6-14 " & to == "Red + Pink       ") ~ TRUE,
    TRUE ~ FALSE
  ))
foo(edges3)
```

## **Potential** multiple comparison

* Had the data been different, you would end up with a different model

* Analysis contingent on data $\rightarrow$ $p$-value argument disappears

$~$

  *"One can no longer argue that, if nothing were going on, that something as extreme as what was observed would occur less than 5% of the time."*

## Q: How to resolve this?

```{r preregistration}
#| out.width: "60%"
f("images/preregistration.png")
```

* Open Science Framework (OSF):

  - https://help.osf.io/article/145-preregistration

* As Predicted:

  - https://aspredicted.org/

## Q: What about null results?

```{r registered-reports}
#| out.width: "70%"
f("images/registered-reports.png")
```

* Training!

  - Experts on preregistration & registered reports within University

## Q: But many hypotheses were not well formed before

:::::: {.columns}
::: {.column width="54%" data-latex="{\textwidth}"}

$$~$$

* Don't treat the finding as definitive truth

* More like exploratory analysis

* Reproduce/replicate it!

  - Experiment, collect data & analyse again

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="44%" data-latex="{\textwidth}"}

```{r xkcd}
#| out.width: "70%"
f("images/replication_crisis.png")
```

Source: https://xkcd.com/3117/

:::
::::::

## Quotes from discussion

* *"... evidence in these research papers is not as strong as stated."*

* *"... watch out for a system that allows any hunch to be ratched up to a level of statistical significance that is then taken as scientific proof."*

* *"... it is ironic that, by way of data-dependent analyses, $p$-values are often used to lend credence to noisy claims based on small samples."*

## Why psychology?

* Small & noisy data

  - Small effect sizes & sample sizes
  - Large measurement errors & high variation

* More common in social sciences

* Much higher threshold in other fields

  - In physics: $p$-values $<3\times10^{-7}$ required

## Some more context

:::::: {.columns}
::: {.column width="49%" data-latex="{\textwidth}"}

\begin{align*}
~&~
\end{align*}

* Original article unpublished

  - Citation count $>1000$

* A similar version in @gl14

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="49%" data-latex="{\textwidth}"}

```{r p-value}
f("images/p-value.jpg")
```

Source: https://www.americanscientist.org/article/the-statistical-crisis-in-science

:::
::::::

## Some more context

```{r gelman}
#| out.width: "50%"
f("images/gelman.jpg")
```

* Fighting bad papers on his blog: https://statmodeling.stat.columbia.edu

```{r himmicane}
#| out.width: "60%"
f("images/himmicane.png")
```

## References