---
title: "Networks $\\times$ Extremes Reading Group"
author: "Clement Lee"
date: "2021-11-11 (Thu)"
output:
  slidy_presentation:
    toc: true
    duration: 30
---

```{r prelim, echo = FALSE, message = FALSE}
library(tidyverse)
library(igraph)
knitr::opts_chunk$set(fig.align = "center", echo = FALSE, out.width = "70%", warning = FALSE)
ggplot2::theme_set(ggplot2::theme_classic(25))
```





## Background
* Directed networks here
* Preferential attachment model:
  - Nodes join the network one by one, bringing in new edges
  - $\Pr($existing node $i$ gets an edge$)~\propto~$ current degree of $i$
  - In-/total degree captures the rich-gets-richer effect





## Simulated network
* Size according to in-degree

```{r layout}
knitr::opts_chunk$set(out.width = "60%")
layout_coord <- function(x) {
  z <- sqrt(x)
  y <- floor(z)
  w <- ifelse(y == z,
    ifelse(y %% 2L == 1L, 1L, 2L),
	ifelse(y %% 2L == 1L,
	  ifelse(x - y ^ 2L <= y, 3L, 4L),
	  ifelse(x - y ^ 2L <= y, 5L, 6L)
	)
  )
  xcoord <- recode(w,
    (y - 1L) / 2L, # case 1
	1L - y / 2L, # case 2
	(y - 1L) / 2L + 1, # case 3
	(y ^ 2L + 3L * (y + 1L) / 2L - x), # case 4
	- y / 2L, # case 5,
	y / 2L - (y + 1) ^ 2L + x # case 6
  )
  ycoord <- recode(w,
    (y - 1L) / 2L, # case 1
	- y / 2L, # case 2
	(y - 1L) / 2L - x + y ^ 2L + 1L, # case 3
	- (y + 1L) / 2L, # case 4
	- y / 2L + x - y ^ 2L - 1L, # case 5
	y / 2L # case 6
  )
  cbind(xcoord, ycoord)
}
```

```{r PA_model, out.width = "60%"}
## BRG model
n0 <- 5 #### beginning number of nodes
g0 <- make_empty_graph(n0, directed = TRUE) %>%
    add_edges(c(1, 2, 2, 3, 2, 4, 3, 4, 4, 5))
V(g0)$label <- labels <- head(LETTERS, n0)
degs <- degree(g0)
edges0 <- igraph::as_data_frame(g0) %>% as.matrix() %>% t() %>% as.integer()
## Now PA model
mu <- 5.0 #### mean number of edges
c2 <- 0.01 #### additive constant
n2 <- 121 #### total number of nodes
g2 <- make_empty_graph(n = n2, directed = TRUE) %>%
    add_edges(edges0)
l2 <- layout_coord(seq(n2))
set.seed(21L)
for (i in (n0 + 1):n2) {
  E2 <- min(rgeom(1, 1.0 / mu) + 1, i - 1) # number of new edges; the plus 1 makes mu meaningful
  p2 <- head(degree(g2, mode = "all"), i - 1) + c2 # "all" leads nicely to power law, "in" doesn't
  nodes2 <- sample(seq(i - 1), E2, prob = p2 / sum(p2))
  edges2 <- as.integer(t(matrix(c(rep(i, E2), nodes2), ncol = 2)))
  g2 <- g2 %>% add_edges(edges2)
  d2 <- degree(g2, mode = "in") + 1 # artificial
  V(g2)$frame.color <- "white"
  V(g2)$color <- "red"
  V(g2)$color[d2 == 0] <- NA
  V(g2)$size <- d2 ^ 0.5 * 3 # for aesthetics, trial-and-error
}
par(mar = rep(0.0, 4))
plot(g2, layout = l2, vertex.label = NA, edge.arrow.size = 0.2, margin = c(0.5, 0.1, 0.1, 0.1), edge.width = 0.2)
```

## Marginal distribution

```{r in_marginal}
n3 <- 1000
g2 <- g2 %>% add_vertices(n3 - n2)
for (i in (n2 + 1):n3) {
  E2 <- min(rgeom(1, 1.0 / mu) + 1, i - 1) # number of new edges; the plus 1 makes mu meaningful
  p2 <- head(degree(g2, mode = "all"), i - 1) + c2 # "all" leads nicely to power law, "in" doesn't
  nodes2 <- sample(seq(i - 1), E2, prob = p2 / sum(p2))
  edges2 <- as.integer(t(matrix(c(rep(i, E2), nodes2), ncol = 2)))
  g2 <- g2 %>% add_edges(edges2)
  d2 <- degree(g2, mode = "in") + 1 # artificial
  V(g2)$frame.color <- "white"
  V(g2)$color <- "red"
  V(g2)$color[d2 == 0] <- NA
  V(g2)$size <- d2 ^ 0.5 * 3 # for aesthetics, trial-and-error
}
df2 <- tibble("in" = degree(g2, mode = "in"), out = degree(g2, mode = "out"))
df2 %>% count(`in`) %>% ggplot() + geom_point(aes(`in`, n)) + scale_x_log10() + scale_y_log10() + labs(x = "in-degree", y = "count")
```





## Why?
* Easy to simulate from and fit to data
* In-degrees of empirical networks seem to follow the "power law"
  - Essentially the (non-generalised) Pareto distribution
  - Some argue otherwise (e.g. Broido and Clauset, 2018)
  
* Power law adequate? Not today's scope





## Making the model more realistic
* $\Pr($existing node $i$ gets an edge$)~\propto~$ current degree of $i$ + $\boldsymbol{\delta}$
  - A positive probability of getting connected even if the node has no in-degre to start with
  - Level the playing field

* Reciprocity parameter $\boldsymbol{\rho}$
  - So far only either $A\rightarrow B$ or $B\rightarrow A$, but not both $\quad\Rightarrow\quad$ reciprocity $=0$
  - Empirical networks usually show high reciprocity
  - E.g. Instagram: if $A$ follows $B$, $B$ is more likely to follow $A$ back
  - $\rho$ affects how the in-degrees & out-degrees are correlated





## Joint distribution
* Same simulated network with 0 reciprocity

```{r in_out_joint}
df2 %>% ggplot() + geom_jitter(aes(`in`, out)) + labs(x = "in-degree", y = "out-degree")
```





## Wang & Resnick (2021)
How does incorporating reciprocity in the preferential attachment model affect the in- & out-degrees?

1. What is their limiting (marginal) distribution?
2. What is their limiting joint distribution?
3. Are they asymptotically independent / dependent?





## Summary
* A relatively restrictive preferential attachment model even with reciprocity
  - Possibly for the theoretical derivations
* Turning the process of preferential attachment into a **multitype branching process with immigration**
* Using results developed for branching process to arrive at the limiting distributions of in- & out-degrees
  - Thus answering the questions they set out





## Model set up
1. Start with node 1 with a self-loop
2. Node $n+1$ joins when there are $n$ nodes in the network
3. With probability $\alpha$, there is an edge from node $n+1$ to an existing node  
    - The node is chosen with probability $~\propto~$ their current in-degree $+~\delta$  
    - With probability $\rho$, reciprocity happens i.e. an edge from the chosen node to $n+1$ is created  
    - With probability $1-\rho$, reciprocity doesn't happen
4. With probability $\gamma(=1-\alpha?)$, there is an edge from an existing node to node $n+1$
    - The node is chosen with probability $~\propto~$ their current out-degree $+~\delta$  
	- With probability $\rho$, reciprocity happens i.e. an edge from $n+1$ to chosen node is created  
	- With probability $1-\rho$, reciprocity doesn't happen





## Converting to branching processes
* An in-degree $~\Leftrightarrow~$ a type 1 particle
* An out-degree $~\Leftrightarrow~$ a type 2 particle
* A node joins $~\Leftrightarrow~$ immigration event
  - Node $n+1$ joins at time $T_{n}$
* Creation of new edge $~\Leftrightarrow~$ death of existing particle & birth of new particle(s)
* While a type 1 particle can give birth to a type 2 particle, births of new type 1 particles must come with it $~\Leftrightarrow~$  
  While an in-degree can lead to a new out-degree, there must be a new in-degree first





## Converting to branching processes (cont'd)
* The existing node that connects with node $n+1$ $~\Leftrightarrow~$ $J_{n+1}$
* Reciprocity occurs when node $n+1$ joins $~\Leftrightarrow~$ $R_{n+1}$
* The evolution of the in- & out-degree of node $k$ $~\Leftrightarrow~$ $\boldsymbol{\xi}_{1,\delta}(\cdot)$
* The allocation of new edge(s) from node $n+1$ depends only on current in- & out-degrees $~\Leftrightarrow~$  
  The branching process is Markovian





## Connecting the two formally
* The collection of the branching processes 
  $\left\{\boldsymbol{\xi}^{*}_\delta(T_n):n\geq0\right\}$
  - Before node $n+1$ arrives, $\boldsymbol{\xi}_{n+1},\delta(\cdot)$ is set to $(0,0)$
* The collection of in- & out-degree sequences $\boldsymbol{D}(n):=\left((D_1^{in}(1), D_1^{out}(1)), \ldots, (D_1^{in}(n), D_1^{out}(n)), (0,0), \ldots\right)$
* Theorem 2: The two have the same distributions
  - Any result for $\left\{\boldsymbol{\xi}^{*}_\delta(T_n):n\geq0\right\}$ can be used for the in- & out-degree sequences





## Result 1 / Theorem 4
* When there are $n$ nodes in the network, the in-degree of node $\omega$, rescaled by $n$ raised to a certain power, converges to a finite random variable $L_\omega$, as $n\rightarrow\infty$.
* Similar for the out-degree
* $n$ raised to a certain power -- power law?
  - This power is a function of the parameters $(\alpha,\gamma,\rho,\delta)$
  - We will come back to this
* Special case: when $\alpha=1$
  - Nodes selected only according to in-degrees
  - Reciprocity only from existing node to new node
  - The rescaled out-degree converges to $\rho~L_\omega$





## Result 2 / Theorem 5
* Joint degree counts
  - Joint distribution of in- & out-degree in a different way
* $N_{m,l}(n)$: the number of nodes with $m$ in-degree & $l$ out-degree when there are $n+1$ nodes in the network
  - $N_{m,l}(n)/n$: Joint empirical pmf of $(m, l)$
* The first line of (3.10) not explicitly simplified, but leads to next result





## Result 3 / Theorem 6
* The in- & out-degrees are jointly regularly varying
* Index $(1+\rho+\delta)/\lambda_1$
  - Echoing theorem 4?
* Special case: when $\alpha=1$, $\lambda_1=1$
  - MRV with index $1+\rho+\delta$
* Asymptotic dependence
  - Question: Even when $\rho=0$?





## Full asymptotic dependence
* The large $(I,O)$ pairs concentrate on the straight line $y=ax$
* Echoed by middle plot of Figure 4.1
* Calculate $O-aI$ and investigate if there's hidden regular variation





## Simulation & Application
* Without modification i.e. fixed reciprocity: middle plot of Figure 4.1
  - Behaviour as expected
* Attempted to randomise $\rho$ to make the model more realistic i.e. closer to Facebook data
  - Left & right plot of Figure 4.1
  - Different user pairs have different degrees of reciprocity





## Some thoughts
* Model quite restrictive & not very realistic
* Asymptotic dependence even if there's no reciprocity
* Only one edge is added at each immigration
  - Presumably derivations are far more challenging when bringing in multiple edges & sampling without replacement
* No connecting between existing nodes
  - They did mention this in extensions



