---
title: "STOR601 - Research Topics"
subtitle: "Network Modelling"
author: "Clement Lee"
date: "2021-01-14 (Thu)"
output:
    beamer_presentation
header-includes:
    \setlength{\leftmargini}{\labelwidth}
    \usetheme[]{lancasterbeamer}
classoption: "aspectratio=169"
---

```{r prelim, include = FALSE, message = FALSE}
library(tidyverse)
library(crandep)
library(igraph)
knitr::opts_chunk$set(out.width = "45%", fig.align = "center", message = FALSE, fig.asp = 0.7, echo = FALSE)
ggplot2::theme_set(ggplot2::theme_bw(40))
```

## Welcome
* Outline of this hour

  - Stochastic block models
  - Modelling degree distribution

# Stochastic block models

```{r sbm_example}
s <- c(25, 30, 35)
K <- length(s)
n <- sum(s)
Y <- matrix(0.0, n, n)
set.seed(1000L)
p1 <- 0.8
p2 <- 0.05
for (i in seq(s[1])) {
    for (j in seq(s[1])) {
        if (i < j) Y[i, j] <- rbernoulli(1, p1)
    }
    for (j in (seq(s[2]) + s[1])) {
        if (i < j) Y[i, j] <- rbernoulli(1, p2)
    }
    for (j in (seq(s[3]) + s[2] + s[1])) {
        if (i < j) Y[i, j] <- rbernoulli(1, p2)
    }
}
for (i in (seq(s[2]) + s[1])) {
    for (j in (seq(s[2]) + s[1])) {
        if (i < j) Y[i, j] <- rbernoulli(1, p1)
    }
    for (j in (seq(s[3]) + s[2] + s[1])) {
        if (i < j) Y[i, j] <- rbernoulli(1, p2)
    }
}
for (i in (seq(s[3]) + s[2] + s[1])) {
    for (j in (seq(s[3]) + s[2] + s[1])) {
        if (i < j) Y[i, j] <- rbernoulli(1, p1)
    }
}
Y1 <- Y + t(Y) # make it sym. then undir.
Y0 <- Y1
diag(Y0) <- NA
g0 <- graph_from_adjacency_matrix(Y1, "undirected")
m <- length(E(g0))
Z <- c(rep(1L, s[1]), rep(2L, s[2]), rep(3L, s[3]))
c0 <- list(membership = Z)
class(c0) <- "communities"
### calculating densities
f <- function(i, j) {round(mean(Y0[Z==i,Z==j], na.rm = T), 3)}
mat0 <- matrix(
    c(f(1,1), f(1,2), f(1,3),
      f(2,1), f(2,2), f(2,3),
      f(3,1), f(3,2), f(3,3)),
    nrow = 3, ncol = 3
)
m1 <- mean(seq(n)[Z == 1])
m2 <- mean(seq(n)[Z == 2])
m3 <- mean(seq(n)[Z == 3])
df0.block <- tribble(~x, ~y, ~z,
    m1, m1, f(1, 1),
	m1, m2, f(1, 2),
	m1, m3, f(1, 3),
	m2, m1, f(2, 1),
	m2, m2, f(2, 2),
	m2, m3, f(2, 3),
	m3, m1, f(3, 1),
	m3, m2, f(3, 2),
	m3, m3, f(3, 3)
)
```

## Example: `r n` nodes, `r m` edges
```{r plot_eg_circle}
plot(g0, vertex.size = 7, vertex.label.cex = 0.8, layout = layout_in_circle)
```

## Visually better
```{r plot_eg_better}
set.seed(1001)
plot(g0, vertex.size = 11, layout = layout_with_fr)
```

## Colouring the groups
```{r plot_eg_colour}
set.seed(1001)
plot(c0, g0, vertex.size = 11, edge.color = "grey50", layout = layout_with_fr)
```

## Observations
* Nodes are tightly connected to each other **within** each group
  - Less connected **between** groups
* More importantly, the connection pattern is **similar** for all nodes within the same group

  - Approx. same number of edges with nodes in the same group
  - Approx. same number of edges with nodes in another group

* I simulated the data & made the communities visible
  - Less clear-cut for real-life data

## Motivation
* Given: a network
  - Undirected or directed
  - There could be node/edge attributes
  - **Not** considered as classified labels though
* To detect:

  - Groups in which nodes are tightly connected (community detection); **and/or**
  - Groups in which the connectivity pattern is similar

* Node clustering / unsupervised learning

## Community detection **algorithms**
* Related to stochastic block models

* Heuristic algorithms
* Quick, based on simple rules
* Comprehensive review by [Fortunato (2010, Physics Reports)](https://www.sciencedirect.com/science/article/abs/pii/S0370157309002841)
* Let's look at an example

## Label propagation algorithm
* [Raghavan, Albert and Kumara (2007, *Physical Review E*)](https://journals.aps.org/pre/abstract/10.1103/PhysRevE.76.036106)

* Initially each node is randomly assigned to one of the $K$ groups
* Each node takes turn to join the group to which the maximum number of neighbours belong
* The iterative process continues until no node changes its group membership anymore

## Issues of community detection algorithms
* Different initial configurations may lead to different results

* Different algorithms give different results
  - Which one is better / best?
* Pre-specifying the number of groups beforehand

## Stochastic block models (SBMs)
* Proper statistical models

  - Theoretically one set of results, once the model is specified 

* Idea: To put nodes that have **similar** properties together
  - The concept of **stochastic equivalence**

## Stochastic equivalence
* Given nodes $p$ and $q$ in the same group

* $\Pr(\text{edge from}~p~\text{to}~r) =
   \Pr(\text{edge from}~q~\text{to}~r)$
    - No matter $r$ is in the same group with $p$ & $q$ or not
	- The two probabilities are independent, **given the group memberships**
* This is the one I used for the example

|       | Group 1  | Group 2  | Group 3  |
|:-----:|:--------:|:--------:|:--------:|
|Group 1|  `r p1`  |  `r p2`  |  `r p2`  |
|Group 2|  `r p2`  |  `r p1`  |  `r p2`  |
|Group 3|  `r p2`  |  `r p2`  |  `r p1`  |

## SBMs do not necessarily achieve community detection
* The model equation says nothing about the probabilities' magnitude

* High number of connections between groups but not within groups

|       | Group 1 | Group 2 | Group 3 |
|:-----:|:-------:|:-------:|:-------:|
|Group 1|   0.03  |   0.06  | **0.80**|
|Group 2| **0.90**|   0.05  |   0.01  |
|Group 3|   0.02  | **0.70**|   0.04  |

## Fitting to real networks
* SBMs *usually* achieve community detection

  - Able to cluster tightly connected nodes

* This means the following are assumed:
  - Stochastic equivalence for tightly connected nodes
  - Degrees of nodes within each group follow the same distribution
  - Quite strong assumptions

* Not quite realistic for actual data
  - Degree heterogeneity
  - Lack of fit

## Illustration
```{r two_stars, fig.width = 8, fig.asp = 0.8, out.width = "40%"}
df0.stars <- bind_rows(
    tibble(from = LETTERS[2:13], to = "A"),
	tibble(from = LETTERS[15:26], to = "N"),
	tibble(from = "A", to = "N")
)
df0.letters <- tibble(name = LETTERS)
g0.stars <- graph_from_data_frame(df0.stars, vertices = df0.letters)
c0.stars <- list(membership = c(rep(1L, 13L), rep(2L, 13L)))
class(c0.stars) <- "communities"
set.seed(22L)
plot(c0.stars, g0.stars)
```
* Community detection achieved but poor fit

## Illustration (cont'd)
```{r two_stars_contd, fig.width = 8, fig.asp = 0.8, out.width = "40%"}
c1.stars <- list(membership = c(1L, rep(2L, 12L), 1L, rep(2L, 12L)))
class(c1.stars) <- "communities"
set.seed(22L)
plot(c1.stars, g0.stars)
```
* Nodes stochastic equivalent but not tightly connected

## Degree-corrected SBM
* [Karrer and Newman (2011, *Physical Review E*)](https://journals.aps.org/pre/abstract/10.1103/PhysRevE.83.016107)

* An extra parameter / latent variable for each node

  - To correct for the heterogeneity in their degrees
  - Nodes within the same group "stochastically equivalent" after correcting

* Community detection + improved fit

## Not the universal fix though
* In some cases, selecting between the original and degree-corrected versions is not clear cut

* Each version reveals a different underlying structure about the same network
* Instead of choosing between the two versions, they could be combined in a weighted / mixture model

## Inference
* What data do we have?
  - The network, represented by e.g. the adjacency matrix

* What quantities do we have to infer?

  - The group memberships of the nodes
  - The block matrix

|       | Group 1  | Group 2  | Group 3  |
|:-----:|:--------:|:--------:|:--------:|
|Group 1|    ??    |    ??    |    ??    |
|Group 2|    ??    |    ??    |    ??    |
|Group 3|    ??    |    ??    |    ??    |

## Using the first example
```{r plot_eg_again}
plot(g0, vertex.size = 7, vertex.label.cex = 0.8, layout = layout_in_circle)
```

## Rearranging the adjacency matrix
* From ... 
```{r plot_adj_reordered, fig.asp = 0.85}
df0 <- tibble(old = seq(n), new = sample(seq(n), n))
df1 <- bind_rows(
  g0 %>% as_data_frame(),
  g0 %>% as_data_frame() %>% rename(from = to, to = from)
) %>%
  left_join(df0, c("from" = "old")) %>%
  rename(from_new = new) %>%
  left_join(df0, c("to" = "old")) %>%
  rename(to_new = new)
df1 %>% ggplot() + geom_point(aes(to_new, from_new)) + scale_y_reverse() + theme_void()
```

## Inference achieved
* ... to ...
```{r plot_adj, fig.asp = 0.85}
df1 %>% ggplot() + geom_point(aes(to, from)) + scale_y_reverse() + theme_void()
```

## Adjacency matrix $\rightarrow$ block matrix
* Densities in each block $=$ point estimates of the parameters
```{r plot_adj_labelled, fig.asp = 0.85}
df1 %>% ggplot() + geom_point(aes(to, from), alpha = 0.5) + scale_y_reverse() + geom_text(aes(x, y, label = z), df0.block, color = "red", size = 20) + theme_void()
```

## Inference issues
* Computational cost of naive inference algorithm usually $O(n^2)$

  - $n=$ number of nodes

* Number of groups $K$

  - Infer it as a parameter?
  - Select it using criteria?

* Label switching issue

  - Common in clustering

* Bayesian approach by [Peixoto (2020, *ArXiv:1705.10225*)](https://arxiv.org/pdf/1705.10225.pdf)

  - Biggest development since the degree-corrected SBM
  - Seems to solve most modelling & inference issues

## Extensions
* Assortative SBMs

* Mixed membership SBMs
* Overlapping SBMs
* Models for temporal / dynamic networks
* With textual information

## Assortative SBMs
* We have seen community detection *algorithms*

* Modifying SBMs specifically for community detection purposes

  - Large probabilities within groups (along the major diagonal)
  - Small probabilities between groups (off-diagonal)
  - Assortativeness

* [Lu and Szymanski (2019, *Scientific Reports*)](https://www.nature.com/articles/s41598-019-49580-5) models the assortativeness

## Mixed membership / overlapping SBMs
* Hard clustering so far

  - Each node belongs to one group only

* Mixed membership models
  - A membership **vector** for each node e.g. $(0.2, 0.5, 0.3)$
  - Hard clustering a special case: a $K$-vector with one 1 & the rest 0's
  - [Airoldi et al. (2008, *JMLR*)](http://www.jmlr.org/papers/v9/airoldi08a.html)

* Overlapping models
  - A node can belong to one or more groups at the same time
  - $2^{K}-1$ choices: $\{1\},\{2\},\{3\},\{1,2\},\{2,3\},\{1,3\},\{1,2,3\}$
  - [Ranciati et al. (2020, *AoAS*)](https://projecteuclid.org/euclid.aoas/1600454877)

## $+$ Temporal / textual information
* Instances of the network at different times

* Textual information associated with:

  - The edges: email exchanges network
  - The nodes: citation network

* Incorporate topic and/or longitudinal modelling
* [Corneli et al. (2018, *Statistics and Computing*)](https://link.springer.com/article/10.1007/s11222-018-9832-4)

## Related methods / models
* Community detection algorithms - we have seen them

* Latent feature models
* Latent space models
* Hypergraph models

## Latent feature models
* $K$ *features* instead of $K$ groups

* Having none, one or multiple features

  - $2^K$ choices (compared to $2^K-1$ for overlapping SBMs)

* [Miller et al. (2009, *NIPS*)](https://papers.nips.cc/paper/2009/hash/437d7d1d97917cd627a34a6a0fb41136-Abstract.html)

## Latent space models
* Instead of being put into groups

* Projected onto an Euclidean space
* Being connected to SBMs by e.g.  
  [Sanna Passino and Heard (2020, *Statistics and Computing*)](https://link.springer.com/article/10.1007/s11222-020-09946-6)

## Hypergraph models
* Consider coauthorship networks

  - Each node represents an author
  - Traditionally: edge between two authors if they have coauthored one or more articles

* How to interpret pairwise edges between nodes 1, 2 and 3?
  - Each pair have collaborated separately?
  - All of them have written one or more articles as a whole?
  - A combination of both?

* There is information lost using SBM

## Hypergraph models (cont'd)
* In a hypergraph:

   - Each pair have collaborated separately: 3 hyperedges, namely $\{1,2\}$, $\{2,3\}$, $\{3,1\}$
   - All of them have written one or more articles as a whole: $\{1,2,3\}$

* [Lunagomez et al. (2017, *JASA*)](https://www.tandfonline.com/doi/full/10.1080/01621459.2016.1141686)

## Transition: my work
* [Lee and Wilkinson (2019, *Applied Network Science*)](https://doi.org/10.1007/s41109-019-0232-2)

  - Review of stochastic block models
  - All references before (& more)

* Networks that are **directed acyclic graphs** (DAGs)

  - Citation network of all the papers in a conference
  - [Dependency network of R packages](https://cran.r-project.org/package=crandep)

## Modifying SBM for DAGs
```{r dag_topo, cache = TRUE}
df0.topo <- cran_dependencies %>%
    filter(type == "imports" & !reverse) %>%
	df_to_graph() %>%
	topo_sort_kahn()
```

* Adjacency matrix can be upper triangular upon rearranging
```{r plot_dag_topo}
df0.depends <- cran_dependencies %>%
    filter(type == "imports" & !reverse) %>%
	inner_join(df0.topo, c("from" = "id")) %>%
	rename(from_num = id_num) %>%
	inner_join(df0.topo, c("to" = "id")) %>%
	rename(to_num = id_num)
df0.depends %>% ggplot() +
    geom_point(aes(to_num, from_num, alpha = 1/to_num), size = 0.1) +
	scale_y_reverse(limits = c(NA, 0)) +
	theme_void() +
	theme(legend.position = "none")
```

## Investigating the degree heterogeneity
* Instead of considering it in the context of SBMs

* How about we look at the degree distribution directly?

# Modelling degree distribution

## What does degree heterogeneity mean?
* Nodes not following the same degree distribution

* There may be underlying reasons for such heterogeneity

  - (Unknown) node covariates
  - But they are not the focus here

* Simply fit a univariate distribution to each kind of degrees:

  - In-degree
  - Out-degree
  - Total degree

## Degrees of R packages
* In-degree seems to follow the **power law**
```{r plot_degrees, fig.show = "hold", fig.asp = c(0.76, 0.64)}
g0.cran <- cran_dependencies %>%
    filter(type == "imports" & !reverse) %>%
	df_to_graph(nodelist = rename(cran_dependencies, name = from))
x0.cran <- degree(g0.cran, mode = "in")
gg0.cran <- tibble(x = x0.cran[x0.cran > 0]) %>%
    count(x) %>%
	ggplot() +
	geom_point(aes(x, n)) +
	scale_x_log10() + scale_y_log10() +
	labs(x = "In-degree", y = "Frequency")
gg0.cran
x1.cran <- degree(g0.cran, mode = "out")
gg1.cran <- tibble(x = x1.cran[x1.cran > 0]) %>%
    count(x) %>%
	ggplot() +
	geom_point(aes(x, n)) +
	scale_x_log10() + scale_y_log10() +
	labs(x = "Out-degree", y = NULL)
gg1.cran
```

## The power law
* In probability terms:

$$ \begin{aligned}
  p(x) &\propto x^{-\alpha}, \qquad x>u\\
  \log p(x) &= -\alpha \log x + c
\end{aligned} $$

* Straight line on the log-log scale

  - $\alpha=-1\times$ slope 

* Related concepts:

  - "The rich gets richer"
  - Preferential attachment
  - 80-20 rule / Pareto principle
  - Scale-free networks

## Preferential attachment model
* [Barabasi and Albert (1999, *Science*)](https://science.sciencemag.org/content/286/5439/509)

* Growing the network by simple rules

  - A new node comes in and connect to existing nodes
  - Probability of connecting to the new node $\propto$ current (in-)degree

* Degree distribution follows the power law with $2<\alpha<3$

  - [Clauset et al. (2009, *SIAM Review*)](https://epubs.siam.org/doi/10.1137/070710111)

* These two papers are themselves embodiment of preferential attachment

## Criticisms of the power law
* Not adequate to describe the degree distribution

  - Usually poorer fit in the right tail, where it is more of interest

* Alternative distributions e.g. log-normal better / sufficient

  - [Stumpf and Porter (2012, *Science*)](https://science.sciencemag.org/content/335/6069/665)
  - [Broido and Clauset (2019, *Nature Communications*)](https://www.nature.com/articles/s41467-019-08746-5)

## Connecting to extremes
* Continuous power law $=$ Pareto distribution

* Pareto distribution $\rightarrow$ Generalised Pareto distribution (GPD)

  - Popular in extreme value theory
  - Can model the tail better / flexibly

* GPD $\rightarrow$ Integer-GPD

  - Discretised
  - [Rohrbeck et al. (2018, *AoAS*)](https://doi.org/10.1214/17-AOAS1081)

* Why don't we use this then?

## Sidetrack: extreme value theory
::: columns
:::: column
* Mixture distribution

* Power law for the left "tail"

  - data $\leq u$

* GPD for the right tail

  - data $> u$
  - More flexible

* $u$ is also inferred
  
::::
:::: column
```{r plot_mix, out.width = "100%"}
u <- 12L
df2.cont <- tibble(x = 1:20, y = dmix(x, 2.0, 0.5, 1.0, u, 0.2, TRUE))
df2.cont %>% ggplot() +
    geom_point(aes(x, y)) +
	geom_segment(aes(x, y, xend = x, yend = 0), alpha = 0.5, lwd = 0.5) +
	geom_vline(xintercept = u, col = 2, lty = 2, lwd = 2) +
	labs(y = "PMF p(x)") +
	annotate("text", 16.5, 0.15, label = "Integer-GPD", size = 10) +
	annotate("text", 7, 0.15, label = "Power law", size = 10)
```
::::
:::

## Good fit
```{r mcmc}
## Data, non-MCMC
raw_data_to_dist <- function(x) {
    ## x is vector of frequencies, degrees, etc.
    table0 <- table(x)
    tibble::tibble(x = as.integer(names(table0)),
                   n = as.integer(table0),
                   S = rev(cumsum(rev(n)) / sum(n)))
}
## Credible intervals from MCMC
lower <- 0.005; upper <- 0.995 ####
mcmc2ci <- function(x, tbl, con, geo, length = 1000L) {
    ## get credible intervals from MCMC output
    ## x is data; tbl is data frame of MCMC values
    tbl <- tbl %>% dplyr::filter(cont == con)
    n <- length(x)
    lxs <- seq(log(min(x)), log(max(x)), length = length)
    xs <- unique(as.integer(round(exp(lxs)))) # need unique values
    N <- nrow(tbl)
    mf <- mS <- matrix(NA, N, length(xs))
    for (i in seq(N)) {
        mf[i,] <- crandep::dmix(xs, tbl$xi1[i], tbl$xi2[i], tbl$sig[i], tbl$u[i], tbl$phi[i], geo = geo)
        mS[i,] <- crandep::Smix(xs, tbl$xi1[i], tbl$xi2[i], tbl$sig[i], tbl$u[i], tbl$phi[i], geo = geo)
    }
    tibble::tibble(x = xs) %>%
        dplyr::mutate(f_cil = apply(mf, 2, quantile, p = lower) * n,
                      f_ciu = apply(mf, 2, quantile, p = upper) * n,
                      S_cil = apply(mS, 2, quantile, p = lower),
                      S_ciu = apply(mS, 2, quantile, p = upper))
}
mcmc.cran <- read_csv(
    "mcmc_cran.csv",
	col_types = cols(cont = col_logical(), .default = col_double())
)
ci.cran <- mcmc2ci(x0.cran[x0.cran > 0], mcmc.cran, TRUE, FALSE)
```

```{r plot_fit, warning = FALSE}
raw_data_to_dist(x0.cran[x0.cran > 0]) %>%
    ggplot() +
	scale_x_log10() +
	scale_y_log10(limits = c(1, NA)) +
	geom_point(aes(x, n)) +
	labs(y = "Frequency", x = "In-degree") +
	geom_line(aes(x, f_cil), ci.cran, col = 2, lty = 5) +
	geom_line(aes(x, f_ciu), ci.cran, col = 2, lty = 5)
```

## Next steps
* Can we modify the preferential attachment model to lead to such in-degree distribution?

* How about modifying an SBM? Or a combination of both?

* Work in progress: [Lee and Eastoe (2020, *ArXiv:2008.03073*)](https://arxiv.org/abs/2008.03073)

## Thank you
* Questions?