---
title: "Good coding practices"
subtitle: "Insights North East Workshop"
author: "Clement Lee (Newcastle University)"
date: "2025-02-04 (Tue)"
output: slidy_presentation
---

## Today's outline

:::::: {.columns}
::: {.column width="49%" data-latex="{\textwidth}"}

1. Introduction

   - The case for coding
   - Quick introduction to R
   - Basic plots 

2. Practical 1

3. Data pipeline via the **tidyverse**

4. Practical 2

5. Write documents & presentations

6. Practical 3

7. Q&A, summary

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="49%" data-latex="{\textwidth}"}

* Today's slides: https://bit.ly/insights-north-east

  - I shall put this in the chat

* Toolkits on https://insightsnortheast.co.uk/data/

:::
::::::

<!-- Running this for the first time, feel free to interrupt and ask questions -->

# 1. Introduction

## About me

:::::: {.columns}
::: {.column width="59%" data-latex="{\textwidth}"}

```{r setup}
#| echo: false
knitr::opts_chunk$set(out.width = "50%", message = FALSE, fig.align = "center")
f <- knitr::include_graphics
```

- Lecturer in Statistics at Newcastle University since August 2022

- Teach statistical analysis using R

- Research on statistical models for (social) networks

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="39%" data-latex="{\textwidth}"}

```{r sophia}
#| echo: false
#| fig.align: "left"
#| out.width: "60%"
f("images/sophia.jpg")
```

:::
::::::

## Experience with outside academia

:::::: {.columns}
::: {.column width="34%" data-latex="{\textwidth}"}

```{r freegle}
#| echo: false
#| out.width: "40%"
f("images/freegle-logo.png")
```

* Case study: https://bit.ly/s4s-freegle

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="64%" data-latex="{\textwidth}"}

```{r freegle-network}
#| out.width: "45%"
#| fig.align: "left"
#| echo: false
f("images/network.png")
```

:::
::::::

## About you?

* Write in the chat

* Experience in working with data, and what are your usual tools?

* Have you coded in R (or other programming languages) before?

## Why coding in the first place?

- Future-proof your analysis

  * for others to replicate / reproduce

  * including your future self

- ChatGPT & others making it easier to start

- Also the stepping stone for machine learning and AI

  * but don't start with something completely unfamiliar

  * garbage in, garbage out

- Version control, CI/CD, and beyond

## Other kinds of output

- "*But I need to make presentations, reports, visualisations, etc.*"

- Most can be done nicely via R

- We will see how this can be done

## Why not Python?

- Python great for big data pipelines

  * Interfacing with the Internet
  * Machine learning at scale

- R is better in some other aspects

  * Standalone analysis
  * The whole data pipeline with `tidyverse`
  * The notebook / literate programming approach

- Understand which one is better / more suited for your tasks

## R

- Free & open source

- Download from https://cran.r-project.org/

- Anyone can contribute:

  * https://cran.r-project.org/web/packages/available_packages_by_name.html

  * Some packages about "business", "politics", etc.

## RStudio

- The best interface for using R

- Download from https://posit.co/download/rstudio-desktop/

- Let's have a look

<!-- give a quick tour of R -->

## Data

```{r read-data}
#| echo: true
library(readxl)
table1 <- read_xlsx("assets/Business Rates properties - November 2024.xlsx")
table1
```

- Source: https://new.newcastle.gov.uk/budget-policies-performance-data/open-data/business-economy/business-rates-data/all-current-business-rates-properties

- A table / spreadsheet in Excel (.xls, .xlsx, .csv, ...) $\approx$ a **data frame** / **tibble** in R

## Some summary statistics

```{r business-summary}
#| eval: true
mean(table1$`rateable value`) # the average of rateable value
sd(table1$`LIA`) # standard deviation of variable LIA
summary(table1$`net charges`) # a range of summaries
```

## A histogram

```{r business-hist}
#| eval: true
hist(table1$`net charges`) # histogram of net charges
```

## Better

```{r business-hist-again}
#| eval: true
hist(table1$`net charges`,
     breaks = 100, # approximately 100 bars
     main = NA, xlab = NA) # don't print the title and the x-axis label yet
title(main = "Histogram of net charges for November 2024",
      xlab = "Net charges (£)")
```

## Dates / times

```{r business-time}
#| eval: true
plot(table1$start, table1$`net charges`)
```

## Better

```{r business-time-again}
#| eval: true
#| warning: false
plot(table1$start, table1$`net charges`, log = "y", # on log scale
     main = NA, xlab = NA, ylab = NA, # don't print title & axis labels yet
     cex = 0.1) # size of the points
title(main = "Net charges over the starting date of businesses",
      xlab = "Starting date", ylab = "Net charges (£)")
```

## Preempting the questions

* What if we have an xls or csv file instead?

* `read_xls()` from the same package `readxl`

* `read.csv()` that comes with R

## Preempting the questions

* What if we want to create new columns (by manipulating some existing columns) and plot them?

* Tedious if we need to

  - create them first,

  - re-save the spreadsheet,

  - and then rerun the analysis

* It is usually the best **not** to modify the raw / original data

## Preempting the questions

* Plots still not perfect

* Huge skew of net charges $\rightarrow$ uninformative histogram; perhaps log scale help?

* Faffy modifications; instead, we will use the package `ggplot2`

## Preparing for the next section

* We can build a data pipeline within R to

  - read the data,

  - manipulate them, and

  - make very nice plots

* Tidyverse (https://www.tidyverse.org/), arguably the biggest development in R **ever**

  - `ggplot2` is part of it

## Questions so far?

* Also, too fast or too slow?

## 2. Practical 1

* Go to https://bit.ly/insights-practical1

* Go through the practical in groups

* Make summaries and plots using your own data or the business rates data

  - Think about what you want to highlight

* Note any challenges / difficulties, or anything you find useful

* Report your findings

# 3. Data pipeline

## The pipe operator

* Continue with the previous data frame `table1`

* Let's say we want to calculate the mean of **the square root of** the net charges

```{r pipe0}
mean(sqrt(table1$`net charges`))
```

* The way the functions are typed is not that the way we think about applying them

* The code below is more **linear**

```{r pipe}
table1$`net charges` |> sqrt() |> mean()
```

## The tidyverse

* Collection of R packages designed for data science

* All packages share an underlying design philosophy, grammar, and data structures

* Installation: only need to do it once (theoretically)

```{r tidyverse-install}
#| eval: false
install.packages("tidyverse")
```

* The above might require a lot more packages to be installed

* After all, it's just a **collection** of packages

## Alternatively

```{r tidyverse-indiv}
#| eval: false
install.packages("dplyr")
install.packages("ggplot2")
install.packages("tidyr")
install.packages("readr")
install.packages("stringr")
install.packages("tibble")
install.packages("purrr")
install.packages("forcats")
```

* Type above commands once and first

* Type below commands every time you need them

  - Only downside: can't type `library(tidyverse)`

```{r tidyverse-indiv-load}
#| eval: true
library(dplyr) # manipulates columns
library(ggplot2) # for plots
library(tidyr) # pivot tables & more
library(readr) # reads tables nicely
library(stringr) # works with text data
library(tibble) # nice tables
library(purrr) # for functional programming
library(forcats) # for "factors"
```

## `dplyr`

* Create / manipulate columns

```{r dplyr}
table2 <- table1 |>
  mutate(date = as.Date(start), ratio = `net charges` / LIA) |>
  select(date, ratio, `rateable value`:`net charges`, "address") # select / reorder columns
table2
```

## `dplyr`

* Filter rows where `DRR1` is not empty (`NA` in R)

* Other uses (not shown):

  - Join other tables

  - Summarise columns

```{r dplyr-filter}
table2 |> filter(!is.na(DRR1))
```

## `tidyr` (+ `dplyr`)

* Pivot tables

  - Combine columns between "rateable value" & "net charges"

  - New columns "name" and "value"

```{r tidyr-pivot}
table3 <- table2 |>
  pivot_longer(`rateable value`:`net charges`) |>
  select(date, ratio, name, value, everything()) |> # a dplyr function
  filter(!is.na(value)) # another dplyr function
table3
```

## `stringr` (+ `dplyr`)

* Extract postcode

* Useful if we want to combine with coordinates for plotting on a map

```{r stringr}
table3 |>
  mutate(postcode = str_extract(address, "NE\\d+ \\d\\w{2}")) |> # within a dplyr function
  select(-address) # postcode suffices for finding the coordinates
```

## `ggplot2`

```{r ggplot2-setup}
#| echo: false
knitr::opts_chunk$set(warning = FALSE)
```

* Histogram of new column `ratio` we have created

* Always the same syntax, as we will see

```{r ggplot2-ratio}
table2 |> ggplot() +
  geom_histogram(aes(ratio))
```

## `ggplot2`

* **Density** of new column `ratio` we have created

* Change the background colour and label size

```{r ggplot2-density}
table2 |> ggplot() +
  geom_density(aes(ratio)) +
  theme_bw(18)
```

## Chaining every together

```{r ggplot2-chain}
table1 |>
  mutate(date = as.Date(start), ratio = `net charges` / LIA) |>
  select(date, ratio, `rateable value`:`net charges`, "address") |>
  ggplot() +
  geom_density(aes(ratio)) +
  theme_bw(18)
```

## Chaining every together

```{r ggplot2-chain-again}
table1 |>
  mutate(date = as.Date(start), ratio = `net charges` / LIA) |>
  select(date, ratio, `rateable value`:`net charges`, "address") |>
  ggplot() +
  geom_point(aes(date, ratio)) +
  theme_bw(18)
```

## What about the problematic net charges?

```{r ggplot2-net-charges}
table1 |>
  ggplot() +
  geom_histogram(aes(`net charges`))
```

## Log scale on the $y$-axis

```{r ggplot2-net-charges2}
table1 |>
  ggplot() +
  geom_histogram(aes(`net charges`)) +
  scale_y_log10()
```

## Polishing the plot

```{r ggplot2-net-charges3}
table1 |>
  ggplot() +
  geom_histogram(aes(`net charges`)) +
  scale_y_log10() +
  scale_x_continuous(labels = scales::comma) +
  theme_bw(18) +
  labs(title = "Histogram of net charges for November 2024", x = "Net charges (£)")
```

## Taking it all the way

```{r ggplot2-net-charges4}
table1 <- read_xlsx("assets/Business Rates properties - November 2024.xlsx")
table1 |>
  ggplot() +
  geom_histogram(aes(`net charges`)) +
  scale_y_log10() +
  scale_x_continuous(labels = scales::comma) +
  theme_bw(18) +
  labs(title = "Histogram of net charges for November 2024", x = "Net charges (£)")
```

## Semi-automation

```{r ggplot2-net-charges5}
month_year <- "November 2024"
table1 <- read_xlsx(paste0("assets/Business Rates properties - ", month_year, ".xlsx"))
table1 |>
  ggplot() +
  geom_histogram(aes(`net charges`)) +
  scale_y_log10() +
  scale_x_continuous(labels = scales::comma) +
  theme_bw(18) +
  labs(title = paste0("Histogram of net charges for ", month_year), x = "Net charges (£)")
```

## Changes in content & title

```{r ggplot2-net-charges6}
month_year <- "August 2024"
table1 <- read_xlsx(paste0("assets/Business Rates properties - ", month_year, ".xlsx"))
table1 |>
  ggplot() +
  geom_histogram(aes(`net charges`)) +
  scale_y_log10() +
  scale_x_continuous(labels = scales::comma) +
  theme_bw(18) +
  labs(title = paste0("Histogram of net charges for ", month_year), x = "Net charges (£)")
```

## As long as format of file name stays the same

```{r ggplot2-net-charges7}
month_year <- "May 2024"
table1 <- read_xlsx(paste0("assets/Business Rates properties - ", month_year, ".xlsx"))
table1 |>
  ggplot() +
  geom_histogram(aes(`net charges`)) +
  scale_y_log10() +
  scale_x_continuous(labels = scales::comma) +
  theme_bw(18) +
  labs(title = paste0("Histogram of net charges for ", month_year), x = "Net charges (£)")
```

## Questions?

## 4. Practical 2

- Install the packages first if you haven't already

- Look at the examples in the slides, and try to do the same for your data

- Look at the package manuals for inspiration: https://www.tidyverse.org/packages/

- If you know what plots you want but not the code, ask

- Note anything useful / challenging, and report your findings

# 5. Write documents & presentations

## What good is it for me?

* Write reports or make presentation slides from time to time

* An imagined typical workflow

  - Do the analysis, create and save the plots

  - Copy the numbers and plots to the word / powerpoint document

  - Discover errors / necessary changes, go back and forth

  - Eventually, some results don't add up

## R Markdown to the rescue

* A script that contains text and code interweave

* Upon a button click:

  - R goes through the code and generates results **on the fly**, then

  - creates the document / slides automatically, and

  - in a format of your choice: pdf / word / HTML / powerpoint

* HTML highly useful

  - Hosting on website

  - Checking the accessibility box

* Testament: I create these materials in this approach

## Workflow part 1

* In RStudio, below **File** in the top left corner, click the small black triangle, then click **R Markdown**

* Enter the title, author, date, and choose the output format (which can always be changed later)

* Click **OK** to open a new file

* Edit between the two lines of triple dashes if necessary

* Grey code **chunks** ringfenced by the two lines of triple backticks

  - Write your R code within

  - Create new one by clicking green square (near top right of the file) and then **R**

* White space is for text that describes your analysis

  - Any blue text will be formatted nicely e.g. section headings, URLs, bold/italics

  - Leave the template text for reference

## Workflow part 2

* Remove the template text when you feel confident

* When ready:

  - Click **Knit** (and save the script file for the first time) and wait;

  - The **Render** panel (bottom left) will show the progress;

  - Once finished, the file may be opened automatically; if not, in the **Files** panel (bottom right), click the file.

* If you want to change the format:

  - Click the small black triangle next to **Knit** and choose the format

  - Same content, different formats at (almost) no cost

  - Complete list here: https://rmarkdown.rstudio.com/lesson-9.html

  - **Caution**: pdf formats (`pdf_document`, `beamer_presentation`) might encounter more issues

## Chunk options

* Towards the top of the file

* The global options

````{verbatim, lang = "markdown"}
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, # print the code or not?
                      message = FALSE, # print messages or not?
                      warning = FALSE, # print warnings or not?
                      include = FALSE) # print results or not?
```
````

* Individual chunk options override the global ones

* The second chunk (with `#|`) looks neater & easier to read

````{verbatim, lang = "markdown"}
```{r setup2, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
table1 |> ggplot() + geom_histogram(aes(`net charges`))
```  


```{r setup3}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#| include: FALSE
table1 |> ggplot() + geom_histogram(aes(`net charges`))
```  
````

* Other options: `fig.cap` for captions of plots

* Full list here: https://yihui.org/knitr/options/

## Self-contained

* Whatever objects you have created or packages loaded in **console** is independent of the R Markdown file

* The code in the Rmd file has to load all necessary packages & create objects within

* Reproducibility

  - Can your colleagues replicate the results?

  - Can you generate the document if the data changes? Related to semi-automation

* Some commands that should not be in an R Markdown file

  - `install.packages(<package name>)` -- `library(<package name>)` suffices

  - `setwd("C:~/home/clement/paths/that/only/exist/on/my/laptop/")`

## Questions?

## 6. Practical 3

- Refer to previous slides (Workflow part 1, Workflow part 2, Chunk options, Self-contained) if needed

- Put the code you have written in Practical 2 in an R Markdown file

- Knit to different formats and see the results

- If in doubt, knit a new file without removing the template, before making changes

# 7. Summary

## Taking stock

* Introduction to R & basic plots

* Data pipeline via the tidyverse (& pipe operator)

  - read the data into R,

  - create / manipulate columns,

  - make plots

* Make documents / presentations using R Markdown

  - Code & text together in one script file

  - Document created upon a button click

## Is this way of making documents useful for me?

:::::: {.columns}
::: {.column width="49%" data-latex="{\textwidth}"}

Cons

- Needs more effort to refine layout

- Needs to set things up & install packages

- Needs all code working before getting the finished product

- Up-front cost of learning the coding before achieving efficiency

:::
::: {.column width="2%" data-latex="{\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
::: {.column width="49%" data-latex="{\textwidth}"}

Pros

- Focus on content, which is separate from format 

- Semi-automation

- Reproducible

- Easy to go on to version control, CI/CD etc. in your workflow

:::
::::::

## "Learning all this is very costly"

* Aggregation of marginal gains (Kaizen)

* Start from the templates, make sure they work, then slowly add your analysis

* Make one plot or calculate one summary at a time

* The manuals (https://www.tidyverse.org/packages/)

* ChatGPT also useful to generate the initial code

## Thank you for listening

* Email: clement.lee@newcastle.ac.uk

* Blog: https://clement-lee.github.io/

* University pages:

  - Me: https://www.ncl.ac.uk/maths-physics/people/profile/clementlee.html

  - NU Solve: https://www.ncl.ac.uk/maths-physics/engagement/nusolve/

* Questions?
